Finally Tagless, Partially Evaluated

### 2008-06-05 tagless/tagless-final.tex
Ken starts to work on some reviewer comments

Our main idea is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using cogen functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

@@

Our solution is \emph{not} Church-encoding the
universal type. The Church encoding of the type~|u| in \S\ref{tagproblem}
requires two continuations; the function |app| in the interpreter above would
have to provide both to the encoding of~|e1|. The continuation
corresponding to the |UB| case of~|u| must either raise an error or
loop. For a well-typed object term, that error continuation is never
invoked, yet it must be supplied. In contrast, our interpreter has no error
continuation at all.

@@

\begin{enumerate}
\item We build interpreters that evaluate (\S\ref{language}),
   compile (or evaluate with staging) (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\ifshort\else
\item We show the first typed, tagless, and statically type-preserving CPS
    transformation with simple types (\S\ref{variations}), which is available for mainstream
    functional languages (ML and Haskell). Type preservation is assured by
    the soundness of our embedding.
\fi
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

To reiterate, we do \emph{not} propose any new language
feature or \ifshort new \else even any new programming \fi technique.
\ifshort
We
\else
Rather, we solve an open problem by a novel combination of simple types
and existing techniques.  More precisely, we
\fi
use features already present in
mainstream functional languages---Hindley-Milner
type system with either an inference\hyp preserving module system or
constructor classes, as realized in ML and Haskell 98---and techniques
which have all appeared in the literature (in particular, \cite{yang-encoding,sumii-hybrid}), to
solve a problem that was stated in the published record as
unsolved and likely unsolvable in ML or Haskell 98
without extensions. The simplicity of our solution and its
use of only mainstream features \ifshort\else are virtues that \fi make it more practical to build typed,
embedded DSLs.
\begin{comment}
We may claim some contribution about type-preserving CPS. Check
related work section in \citep{Guillemette-Monier-PLPV}, especially
check the work of Shao on type-preserving CPS in Flint. The PLPV paper
in related work shows other tasks, including closure conversion, which
we may tackle in our approach. We may be able to write a
type-preserving, assured compiler, whose properties and assured by HM.
\end{comment}

@@

To embed the same object language in (Meta)OCaml, we replace the \ifshort type
class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |'c|, an
\emph{environment classifier} \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{S:compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

@@

Inserting brackets and escapes appropriately into the
evaluator~|R| above yields the simple compiler~|C|
\ifshort in Fig.~\ref{fig:interpreter-C}(a).\else below.
\UseVerbatim{5a}
\fi
\ifshort
\begin{figure}[t]
(a) \BUseVerbatim{5a}

@@

%------------------------------------------------------------------------
% This is the version for the APLAS paper. No Haskell
\ifshort
\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of three
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
a dynamic (future-stage) term (using\ifshort\else\ the
compiler\fi~|C|). To
distinguish between static and dynamic terms, we might try to define
|repr| in the partial evaluator as
\texttt{type ('c,'dv)\,repr 
	= S0 of ('c,'dv)\,R.repr~}\Verb+|+\texttt{~D0 of ('c,'dv)\,C.repr}.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|. We use |C.int| to
convert from the static term |('c,int) R.repr|, which is just |int|,
to the dynamic term.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  Suppose that the first argument to |if_| is a dynamic term
(of type |('c,bool) C.repr|), the second a static term 
(of type |('c,'a) R.repr|), and the third a
dynamic term (of type |('c,'a) C.repr|). We then need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |'a -> ('c,'a) C.repr|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}.

Our |Symantics| only includes separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus, compiling polymorphic lift requires intensional type
analysis.  To avoid needing polymorphic lift, we turn to
Asai's technique \citep{asai-binding-time,sumii-hybrid}:
build a dynamic term
alongside every static term.

We switch to the data type
\texttt{type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always available, we
can now define the polymorphic function
|let abstr1 (P1 (_,dyn)) = dyn| of type
\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code3}
module P1 : Symantics = struct
  let int (x:int) = P1 (Some (R.int x), C.int x)
  let add e1 e2 = match (e1,e2) with
    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
  let if_ = function
    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
                                   (fun () -> abstr1 (et ()))
                                   (fun () -> abstr1 (ee ())))
\end{code3}
However, we stumble on functions.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
}|('c,'a->'a) P1.repr|, consists of a dynamic part 
(type |('c,'a->'a) C.repr|) and
maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1.repr| data structure: the representation
for a function type |'a->'b| should be one of
\begin{code3}
S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
\end{code3}
unlike |P1.repr| of |int| or |bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type. In Haskell, typecase
can be equivalently implemented either with GADTs or with
type-class functional dependencies
\citep{oliveira-typecase}. The accompanying code shows both
approaches, neither portable to OCaml. In addition,
the problem of nonexhaustive pattern\hyp matching reappears in
the GADT approach because GHC 6.6.1 cannot see that a particular
type of a GADT value precludes certain constructors.
Thus GADTs fail to
make it \emph{syntactically} apparent that pattern matching is exhaustive.

Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begingroup\sloppy
\texttt{type ('c,'dv) repr = P1 of (repr\_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code3}
repr_pe ('c,int) = ('c,int); repr_pe ('c,bool) = ('c,bool)
repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
\end{code3}
\endgroup\noindent
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. 
We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
\begin{figure}
\begin{floatrule}
\smallskip
\begin{code2}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
            -> ('c,'x,'da -> 'db) repr
  val app : ('c,'x,'da -> 'db) repr
            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
                                'da -> 'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr
            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
\end{code2}
\medskip
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
  \texttt{mul}, \texttt{leq} are elided)\fi}
\label{fig:ocaml}
\end{figure}

\begin{figure}
\begin{floatrule}
\begin{code2}
module P = struct
  type ('c,'sv,'dv) repr = {st: 'sv option; dy: ('c,'dv) code}
  let abstr {dy = x} = x    let pdyn x = {st = None; dy = x}

  let int  (x:int)  = {st = Some (R.int x);  dy = C.int x}
  let add e1 e2 = match e1, e2 with
  | {st = Some 0}, e | e, {st = Some 0} -> e
  | {st = Some m}, {st = Some n} -> int (R.add m n)
  | _ -> pdyn (C.add (abstr e1) (abstr e2))
  let if_ eb et ee = match eb with
  | {st = Some b} -> if b then et () else ee ()
  | _ -> pdyn (C.if_ (abstr eb) (fun () -> abstr (et ()))
                                (fun () -> abstr (ee ())))
  let lam f = {st = Some f; dy = C.lam (fun x -> abstr (f (pdyn x)))}
  let app ef ea = match ef with {st = Some f} -> f ea
                  | _ -> pdyn (C.app (abstr ef) (abstr ea)) end
\end{code2}
\medskip
\end{floatrule}
\caption{Our partial evaluator (\ifshort \texttt{bool}, \texttt{mul},
  \texttt{leq} and \texttt{fix} \else \texttt{mul} and \texttt{leq} \fi
  are elided)}
\label{fig:pe}
\end{figure}

The interpreters |R|, |L| and~|C| above only use the old
type arguments |'c| and~|'dv|, which are treated by the new signature
in the same way.  Hence, all that needs to change in these interpreters
to match the new signature is to add a phantom type
argument~|'sv| to~|repr|.
For example, the compiler |C| now begins
\ifshort
\texttt{module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code}
\else
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
\end{code}
\fi
with the rest the same.
In contrast, the partial evaluator~|P| relies on the type argument |'sv|.

To illustrate how to add optimizations, we improve |add| (and |mul|,
elided) to simplify the generated code using the monoid (and ring)
structure of~|int|: not only is addition performed statically
(using~|R|) when both operands are statically known, but it is
eliminated when one operand is statically~$0$; similarly for
multiplication by~$0$ or~$1$.  
\ifshort
Such optimizations can be quite effective
\else
Such algebraic simplifications are easy
to abstract over the specific domain (such as monoid or ring) where they
apply.  These simplifications and abstractions help a lot
\fi
in a large language with more base types and primitive operations.
\ifshort\else
Incidentally, the code actually contains a more general implementation
mechanism for such features, inspired in part by previous work in
generative linear algebra~\citep{CaretteKiselyov05}.
\fi

Any partial evaluator must decide how much to unfold recursion
statically: unfolding too little can degrade the residual code, whereas
unfolding too much risks nontermination.  Our partial evaluator is no
exception, because our object language includes |fix|.  The code in
Figure~\ref{fig:pe} takes the na\"\i ve approach of ``going all the
way'', that is, whenever the 
argument is static, we unfold |fix| rather than residualize it.
In the accompanying source code is a conservative
alternative |P.fix| that unfolds recursion only once, then residualizes.
Many sophisticated approaches have been developed to decide how much to unfold
\citep{jones-partial}, but this issue is orthogonal to our presentation.
\ifshort\else
A separate concern in our treatment of |fix| is possible code bloat in
the residual program, which calls for let-insertion
\citep{SwadiTahaKiselyovPasalic2006}.
\fi

Given this implementation of~|P|, our running example
\ifshort
\texttt{let module E\,=\,EX(P) in E.test1 ()}
\else
\begin{code}
let module E = EX(P) in E.test1 ()
\end{code}
\fi
evaluates to
\ifshort
|{P.st = Some true; P.dy = .<true>.}|
\else
\begin{code}
{P.st = Some true; P.dy = .<true>.}
\end{code}
\fi
of type |('a, bool, bool) P.repr|.  Unlike with~|C| in~\S\ref{S:compiler},
a $\beta$-reduction has been statically performed to yield |true|.  More
interestingly, whereas |testpowfix7| compiles to a code value with many
$\beta$-redexes in~\S\ref{S:compiler}, the partial evaluation
\ifshort
\texttt{let module E = EX(P) in E.testpowfix7}
\else
\begin{code}
let module E = EX(P) in E.testpowfix7
\end{code}
\fi
gives the desired result
\ifshort\vspace*{-0.7em}\fi
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x -> x * (x * (x * (x * (x * (x * x)))))>.}
\end{code}
\ifshort\vspace*{-0.7em}\fi

%------------------------------------------------------------------------
% This is the longer version of the section
\else

@@

Our |Symantics| only includes separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus, compiling polymorphic lift requires intensional type
analysis.  To avoid needing polymorphic lift, we turn to
Asai's technique \citep{asai-binding-time,sumii-hybrid}:
build a dynamic term alongside every static term.

@@

The object language generally inherits the evaluation strategy from
the metalanguage---call-by-value (CBV) in OCaml, call-by-name (CBN) in
Haskell.  To represent a CBN object language in a CBV metalanguage,
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}
introduce CPS to make the evaluation strategy of a definitional
interpreter indifferent to that of the metalanguage. To achieve the same
indifference in the typed setting, we build a CBN CPS interpreter for
our object language in OCaml.

@@

We can modify a CBN or CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state. As Figure~\ref{fig:state} shows, we extend our object language
with three imperative features.
\begin{enumerate}
    \item ``$\deref$'' gets the current state;
    \item ``$\set e$'' sets the state to the value of~$e$ and returns
        the previous value of the state;
    \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
        before~$e_2$ even if $e_2$ does not use~$x$.
\end{enumerate}
If $x$ does not appear in~$e_2$, then ``$\lapp{e_1}{x}e_2$'' is same as
the more familiar sequencing form ``$e_1;e_2$''.
We can embed this extended object language into OCaml by extending the
|Symantics| signature in Figure~\ref{fig:ocaml}.
\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states       (* static version of the state *)
  val lapp : (('c,'sa,'da) repr as 'x) -> ('x -> 'y)
             -> (('c,'sb,'db) repr as 'y)
  val deref : unit -> ('c, 'c states, state) repr
  val set   : (('c, 'c states, state) repr as 'x) -> 'x
end
\end{code}
In HOAS, we write $\lapp{e_1}{x}e_2$ as |lapp e1 (fun x -> e2)|;
the type of |lapp| is that of function application with
the two arguments swapped.  We can encode the term
``$\lapp{\deref}{x} (\set 2;\, x+\deref)$''
as the OCaml functor
\begin{code}
module EXSI_INT(S: SymSI
  with type state = int and type 'c states = int) = struct open S
  let test1 () = lapp (deref ()) (fun x -> 
                  lapp (set (int 2)) (fun _ -> add x (deref ())))
end
\end{code}
The accompanying source code shows several more tests, including
a test for higher-order state and a power function that uses state
as the accumulator.

@@

At the term level, we also avoid converting between static and dynamic
terms by building them in parallel, using \citearound{'s
method}\citet{asai-binding-time}.
This method type-checks in Hindley-Milner once we
deforest the object term representation.  Put another way, we
manually apply type-level partial evaluation to our type
functions (see \S\ref{S:PE-solution}) to obtain simpler types 
acceptable to MetaOCaml.

\Citet{sumii-hybrid} also use Asai's method,
to combine online
and offline partial evaluation.
They predate us in deforesting the object term representation to enable tagless
partial evaluation.
We strive for modularity by reusing interpreters for individual stages
\citep{SperberThiemann:TwoForOne}: our partial evaluator~|P|
reuses our tagless evaluator~|R| and tagless compiler~|C|, so it
is patent that the \emph{output} of~|P| never gets
stuck.  It would be interesting
to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
in the same manner.

>>>>

Our main idea is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using compiler-generator (cogen) functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

@@

Unlike \citearound{'s HOAS-based term representation}\citet{JFP-Mogensen},
our solution is \emph{not} Church-encoding the
universal type. The Church encoding of the type~|u| in \S\ref{tagproblem}
requires two continuations; the function |app| in the interpreter above would
have to provide both to the encoding of~|e1|. The continuation
corresponding to the |UB| case of~|u| must either raise an error or
loop. For a well-typed object term, that error continuation is never
invoked, yet it must be supplied. In contrast, our interpreter has no error
continuation at all.

@@

\begin{enumerate}
\item We build interpreters that evaluate (\S\ref{language}),
   compile (or evaluate with staging) (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\ifshort\else
\item We show the first typed, tagless, and statically type-preserving CPS
    transformation with simple types (\S\ref{variations}), which is available for mainstream
    functional languages (ML and Haskell). Type preservation is assured by
    the soundness of our embedding.
\fi
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

To reiterate, we do \emph{not} propose any new language
feature or \ifshort new \else even any new programming \fi technique.
\ifshort
We
\else
Rather, we
\fi
solve a problem that was stated in the published record as open and
likely unsolvable in ML or Haskell 98 without extensions, by a novel
combination of simple types and techniques already described in the
literature that use features present in mainstream functional languages.
In particular, we follow \citearound{'s encoding of type-indexed
values}\citet{yang-encoding}, \citearound{'s construction of dynamic
terms alongside static terms}\citet{asai-binding-time}, and
\citearound{'s deforestation of syntax
constructors}\citet{sumii-hybrid}.  These techniques require just
a Hindley-Milner type system with either a module system that supports
functors or constructor classes, as realized in all variants of ML and
Haskell.
The simplicity of our solution and its
use of only mainstream features \ifshort\else are virtues that \fi make it more practical to build typed,
embedded DSLs.
\begin{comment}
We may claim some contribution about type-preserving CPS. Check
related work section in \citep{Guillemette-Monier-PLPV}, especially
check the work of Shao on type-preserving CPS in Flint. The PLPV paper
in related work shows other tasks, including closure conversion, which
we may tackle in our approach. We may be able to write a
type-preserving, assured compiler, whose properties and assured by HM.
\end{comment}

@@

To embed the same object language in (Meta)OCaml, we replace the \ifshort type
class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |'c|, an
\emph{environment classifier} \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{S:compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.
We shorten some of the types using OCaml's |as| syntax.

@@

To turn the evaluator~|R| into a simple compiler, we bracket the
computation on values to be performed at run time, then escape the code
generation from terms to be performed at compile time.  Adding these
stage annotations yields the compiler~|C|
\ifshort in Fig.~\ref{fig:interpreter-C}(a).\else below.
\UseVerbatim{5a}
\fi
\ifshort
\begin{figure}[t]
(a) \BUseVerbatim{5a}

@@

%------------------------------------------------------------------------
% This is the version for the APLAS paper. No Haskell
\ifshort
\input{tagless-final-aplas}
%------------------------------------------------------------------------
% This is the longer version of the section
\else

@@

Our |Symantics| only includes separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus, compiling polymorphic lift requires intensional type
analysis.  To avoid needing polymorphic lift, we turn to
Asai's technique \citep{asai-binding-time,sumii-hybrid} of
building a dynamic term alongside every static term.

@@

The object language generally inherits the evaluation strategy from
the metalanguage---call-by-value (CBV) in OCaml, call-by-name (CBN) in
Haskell.\footnote{To be more precise, most Haskell implementations
use call-by-need, which is observationally equivalent to call-by-name
because sharing is not observable \citep{ariola-call-by-need-popl}.}
To represent a CBN object language in a CBV metalanguage,
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}
introduce CPS to make the evaluation strategy of a definitional
interpreter indifferent to that of the metalanguage. To achieve the same
indifference in the typed setting, we build a CBN CPS interpreter for
our object language in OCaml.

@@

We can modify a CBN or CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state. As Figure~\ref{fig:state} shows, we extend our object language
with three imperative features.
\begin{enumerate}
    \item ``$\deref$'' gets the current state;
    \item ``$\set e$'' sets the state to the value of~$e$ and returns
        the previous value of the state;
    \item the let-form ``$\lapp{e_1}{x}e_2$'' evaluates $e_1$
        before~$e_2$ even if $e_2$ does not use~$x$.
\end{enumerate}
If $x$ does not appear in~$e_2$, then ``$\lapp{e_1}{x}e_2$'' is same as
the more familiar sequencing form ``$e_1;e_2$''.
We can embed this extended object language into OCaml by extending the
|Symantics| signature in Figure~\ref{fig:ocaml}.
\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states       (* static version of the state *)
  val lapp : (('c,'sa,'da) repr as 'x) -> ('x -> 'y)
             -> (('c,'sb,'db) repr as 'y)
  val deref : unit -> ('c, 'c states, state) repr
  val set   : (('c, 'c states, state) repr as 'x) -> 'x
end
\end{code}
In HOAS\@, we write the term ``$\lapp{e_1}{x}e_2$'' as |lapp e1 (fun x -> e2)|;
the type of |lapp| is that of function application with
the two arguments swapped.  We can encode the term
``$\lapp{\deref}{x} (\set 2;\, x+\deref)$''
as the OCaml functor
\begin{code}
module EXSI_INT(S: SymSI
  with type state = int and type 'c states = int) = struct open S
  let test1 () = lapp (deref ()) (fun x -> 
                  lapp (set (int 2)) (fun _ -> add x (deref ())))
end
\end{code}
The accompanying source code shows several more tests, including
a test for higher-order state and a power function that uses state
as the accumulator.
Because the |SymSI| signature extends |Symantics|, any encoding of
a term in the pure object language (that is, any functor that takes
a |Symantics| module as argument) can also be used as a term in the
extended object language (for example, applied to an implementation of
|SymSI|).

@@

At the term level, we also avoid converting between static and dynamic
terms by building them in parallel, as \citet{asai-binding-time} does
in his partial evaluator.
This method type-checks in Hindley-Milner once we
deforest the object term representation.  Put another way, we
manually apply type-level partial evaluation to our type
functions (see \S\ref{S:PE-solution}) to obtain simpler types 
acceptable to MetaOCaml.

\Citets{JFP-Mogensen} self-reducer for the untyped
$\lambda$-calculus preceded Asai's partial evaluator in constructing
static and dynamic terms in parallel.  However, this self-reducer
constructs a static term for every object term, even a bound variable,
so (in terms of Fig.~\ref{fig:pe}) it moves some work from |app|
to~|pdyn|.  In contrast, we follow Asai and construct the static term
optionally, so as to perform type-directed lifting without requiring the
encoding of an object term to juggle explicit type indices.

\Citet{sumii-hybrid} also use Asai's construction,
to combine online
and offline partial evaluation.
They predate us in deforesting the object term representation to enable tagless
partial evaluation.
We strive for modularity by reusing interpreters for individual stages
\citep{SperberThiemann:TwoForOne}: our partial evaluator~|P|
reuses our tagless evaluator~|R| and tagless compiler~|C|, so it
is patent that the \emph{output} of~|P| never gets
stuck.  It would be interesting
to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
in the same manner.
